---
title: "Visualization of the monthly maps 2020 to 2022"
author: ' Ousmane Diao'
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', cache = TRUE)
options(scipen = 999)
```

```{r}
rm(list=ls())
#gc()
```

```{r libraries}
library(sf)
library(snakecase)
library(lubridate)
library(janitor)
library(tidyverse)
library(viridis)
library(RColorBrewer)
library(ggpubr) # For combining multiple plots
pkg<-c("raster","INLA","TMB","sparseMVN","Matrix","rgdal","gdistance",
       "malariaAtlas","sp","rgeos","foreach","doParallel")
for (i in 1: length(pkg)){library(pkg[[i]],character.only = TRUE)}


library(exactextractr)
library(tidyterra)
```

# Introduction
We are presenting the output from the model with Senegal data. This model is a Baysian framework to predict the malaria cases in pixel level using routine surveillance data from health facilities and considering environmentales covariates, population density, catchment population and treatment seeking probability. 
```{r}
# Load data
outputpath <- "/home/odiao/Output_11_20_2024/"

coordinates <- read_csv("/mnt/efs/stratification/CountryFolders/Senegal//Projects/1.0 Risk Map/2022/01 Raw data/Senegal data for MAP/Org Unit - Corrected Coordinates2.csv") %>% 
  dplyr::select(adm1 = `Org Unit Name Level 2`, adm2 = `Org Unit Name Level 3`,
         hf = `Org Unit Short Name`, lat = Latitude, long = Longitude) %>% 
  mutate(adm1 = snakecase::to_snake_case(adm1),
         adm2 = snakecase::to_snake_case(adm2),
         hf = snakecase::to_snake_case(hf),
         unique_id = paste("sen",row_number(), sep="_"))

bnd <- st_read("/mnt/efs/stratification/CountryFolders/Senegal/Common files/Shapefiles/Shapes files 79 Districts VF/Shapes files 79 Districts VF/districtsanitaire_79_clean.shp")
```

# Analyses

```{r}
data_2020 <- readxl::read_xlsx("/mnt/efs/stratification/CountryFolders/Senegal//Projects/1.0 Risk Map/2022/01 Raw data/PNLPdata/Extraction Data Paludisme DS FS DHIS2 2020_2021_2022.xlsx", sheet = 1)
data_2021 <- readxl::read_xlsx("/mnt/efs/stratification/CountryFolders/Senegal//Projects/1.0 Risk Map/2022/01 Raw data/PNLPdata/Extraction Data Paludisme DS FS DHIS2 2020_2021_2022.xlsx", sheet = 2)
data_2022 <- readxl::read_xlsx("/mnt/efs/stratification/CountryFolders/Senegal//Projects/1.0 Risk Map/2022/01 Raw data/PNLPdata/Extraction Data Paludisme DS FS DHIS2 2020_2021_2022.xlsx", sheet = 3)

```


```{r}
###########################################
# Step 1: Read in new case data
###########################################

# Inspect sample rows:
# head(coordinates)

data_2020_2022 <- bind_rows(data_2020, data_2021, data_2022) %>%
  janitor::clean_names() %>%
  separate(mois, into = c('monthname', 'year'), sep = " ") %>%
  mutate(month = factor(monthname, 
                        levels = c("Janvier","Février","Mars","Avril","Mai","Juin","Juillet","Août","Septembre","Octobre","Novembre","Décembre")),
         month = as.numeric(month),
         year = as.numeric(year),
         year.month= paste(year, month, sep = ".")) %>%
  mutate(adm1 = snakecase::to_snake_case(regions),
         adm2 = snakecase::to_snake_case(districts),
         hf = snakecase::to_snake_case(formation_sanitaire),
         conf_u5 = cas_confirmes_feminin_0_4_ans + cas_confirmes_masculin_0_4_ans,
         conf_ov5 = cas_confirmes_feminin_5_ans_excluant_les_femmes_enceintes + cas_confirmes_masculin_5_ans_excluant_les_femmes_enceintes,
         pres_all = NA,
         pres_u5 = NA,
         pres_ov5 = NA,
         pres_preg = NA) %>%
  dplyr::select(adm1, adm2, hf, month, year, year.month, 
                allout_all = total_consultants_toutes_affections_confondues,
                allout_ov5 = consultants_toutes_affections_confondues_5_ans_excluant_les_femmes_enceintes,
                allout_preg = consultants_toutes_affections_confondues_femme_enceinte,
                allout_u5 = consultants_toutes_affections_confondues_0_4_ans,
                conf_all = total_cas_confirmes, 
                conf_ov5,
                conf_preg = cas_confirmes_femme_enceinte,
                conf_u5,
                pres_all, pres_ov5, pres_preg, pres_u5,
                test_all = total_tests_tdr_realises,
                test_ov5 = tests_tdr_realises_5_ans_excluant_les_femmes_enceintes,
                test_preg = tests_tdr_realises_femme_enceinte,
                test_u5 = tests_tdr_realises_0_4_ans,
                conf_f_all = total_cas_confirmes_feminin,
                conf_f_ov5 = cas_confirmes_feminin_5_ans_excluant_les_femmes_enceintes,
                conf_f_u5 = cas_confirmes_feminin_0_4_ans,
                conf_m_all = total_cas_confirmes_masculin,
                conf_m_ov5 = cas_confirmes_masculin_5_ans_excluant_les_femmes_enceintes,
                conf_m_u5 = cas_confirmes_masculin_0_4_ans,
                susp_all = total_cas_suspect,
                susp_ov5 = cas_suspect_5_ans_excluant_les_femmes_enceintes,
                susp_preg = cas_suspect_femme_enceinte,
                susp_u5 = cas_suspect_0_4_ans) %>%
  left_join(coordinates, by = c("adm1","adm2","hf"))
```

```{r}
###########################################
# Step 3: Summarize coverage of coordinates
###########################################
# Count total unique facilities
facilities_total <- data_2020_2022 %>%
  distinct(adm1, adm2, hf) %>%
  nrow()

# Count how many have coordinates
facilities_with_coords <- data_2020_2022 %>%
  distinct(adm1, adm2, hf, lat, long) %>%
  filter(!is.na(lat)) %>%
  nrow()

# Facilities without coordinates
facilities_without_coords <- facilities_total - facilities_with_coords

# Print summary:
cat("Total number Health facilities: ", facilities_total, "\n")
cat("Total number of facilities with valide coordinates: ", facilities_with_coords, "\n")
cat("Total number of facilities without coordinates: ", facilities_without_coords, "\n")
```


### Population denstiy map
```{r}
pop <- rast("/mnt/efs/stratification/CountryFolders/Senegal/Common files/Population/SEN_population_v1_0_gridded_1km.tif")%>% 
  crop(bnd) %>% mask(bnd) #without: %>% crop(bnd) %>% mask(bnd)

ggplot()+ # with: bnd
  geom_sf(fill = 'grey20')+
  geom_spatraster(data = pop,aes(fill = layer), na.rm = TRUE)+
  scale_fill_distiller(palette = "RdYlGn",na.value = 'transparent',trans="log10",
                       breaks = scales::trans_breaks("log10", function(x) 10^x),
                       labels = scales::trans_format("log10", scales::math_format(10^.x)))+
  coord_sf(crs = 4326)+
  theme_void()+
  labs(title = "Population - GRID3 2020",
    fill = "Pop",size = 18)+
  theme(#legend.key.width = unit(2, "cm"), legend.position = 'bottom'
    plot.title = element_text(size = 18, face = "bold"),   # Title size
    plot.caption = element_text(size = 16),               # Caption size
    legend.text = element_text(size = 16),                # Legend text size
    legend.title = element_text(size = 18, face = "bold"), # Legend title size
    # Adjust legend position and key size
    legend.key.width = unit(1.5, "cm"), 
    legend.key.height = unit(0.5, "cm"),
    legend.position = 'bottom')
```


### Treatment seeking propability map
```{r TS}
ts <- rast("/mnt/efs/stratification/CountryFolders/Senegal/Projects/1.0 Risk Map/2022/03 Additional files (optional)/Treatment seeking/ts_senegal_mean0_2023-03-29.tif") %>% 
  crop(bnd) %>% mask(bnd)

ggplot()+
  geom_spatraster(data = ts,aes(fill = layer), na.rm = TRUE)+
  scale_fill_viridis(option = "D",na.value = 'transparent')+
  coord_sf(crs = 4326)+
  theme_void()+
  labs(title = "Treatment seeking probability",
       fill = "TS prob",
       #caption = "dots represent a geolocated health facility"
       )+
  theme(#legend.key.width = unit(2, "cm"), legend.position = 'bottom'
    plot.title = element_text(size = 18, face = "bold"),   # Title size
    plot.caption = element_text(size = 18),               # Caption size
    legend.text = element_text(size = 18),                # Legend text size
    legend.title = element_text(size = 18, face = "bold"), # Legend title size
    # Adjust legend position and key size
    legend.key.width = unit(1.5, "cm"), 
    legend.key.height = unit(0.5, "cm"),
    legend.position = 'bottom')
#ggsave(paste0(outputpath, "ts_HF_OD.png"))
```


### Travel time minutes map
```{r TT}
tt <- rast("/mnt/efs/stratification/CountryFolders/Senegal/Projects/1.0 Risk Map/2022/03 Additional files (optional)/Treatment seeking/Time_to_HF/2023-03-29/HF.accessibility.tif") %>% 
  crop(bnd) %>% mask(bnd)

ggplot()+
  geom_spatraster(data = tt,aes(fill = layer), na.rm = TRUE)+
  geom_point(data = coordinates, mapping = aes(x = long, y = lat), size = 0.6, col = 'red3', pch=21, show.legend = TRUE)+
  scale_fill_viridis(option = "D",na.value = 'transparent' )+
  coord_sf(crs = 4326)+
  theme_void()+
  labs(title = "Travel time (in minutes) to nearest HF",
    fill = "Tt min",
       #caption = "dots represent a geolocated health facility",size = 18
    )+
  #theme(legend.key.width = unit(2, "cm"), legend.position = 'bottom')
  theme(# Increase title, caption, and legend text size
    plot.title = element_text(size = 18, face = "bold"),   # Title size
    plot.caption = element_text(size = 18),               # Caption size
    legend.text = element_text(size = 18),                # Legend text size
    legend.title = element_text(size = 18, face = "bold"), # Legend title size
    # Adjust legend position and key size
    legend.key.width = unit(1.5, "cm"), 
    legend.key.height = unit(0.5, "cm"),
    legend.position = 'bottom')
 
#ggsave(paste0(outputpath, "tt_HF_OD.png"))
```



```{r input-data}
# Suppress progress bar
options(rasterProgress = 'none')
#Read the tif files contenant the predicted median values
# List raster files
raster_files <- list.files(outputpath, 
  pattern = 'posterior_median_cases_total_per_1000_PYO.tif$', full.names = TRUE)


pop_dat <- read_csv("/mnt/efs/stratification/CountryFolders/Senegal//Projects/1.0 Risk Map/2022/01 Raw data/PNLPdata/Population DS 2015 _ 2023.csv") %>% 
  mutate(adm1 = str_to_title(RM),
         adm2 = str_to_title(Districts)
  ) %>% 
  filter(!is.na(RM)) %>% 
  mutate(adm2 = case_when(adm2 == "Mbao" ~ "MBAO",
                          TRUE ~ as.character(adm2))) %>% 
  dplyr::select(-RM, -Districts) %>% 
  pivot_longer("Populations \n2015":"Populations \n2023", names_to = "year", values_to = "population") %>% 
  mutate(year = str_remove(year, "Populations \n") %>% as.numeric()) %>% 
  filter(year >= 2020 & year < 2023)
```


### Monthly maps in 2020
```{r Incidence maps 2020, fig.width=14, fig.height=8}
pred_inc_ras_2020 <- rast(raster_files[1:12])
names(pred_inc_ras_2020) <- c("pred_01", "pred_02","pred_03","pred_04",
                         "pred_05","pred_06","pred_07","pred_08",
                         "pred_09","pred_10", "pred_11","pred_12")
#plot(pred_inc_ras)
#summary(pred_inc_ras) #This can help to define breaks

pred_inc_2020 <- terra::as.data.frame(pred_inc_ras_2020, xy=TRUE) %>% 
            pivot_longer(cols=names(pred_inc_ras_2020), names_to = "month", values_to = "inc") %>% 
            mutate(month = str_remove(month, 'pred_'),
                   month = as.numeric(month)) %>% 
            mutate(inc_cut = cut(inc, 
                                 breaks = c(0,3, 5, 7, 10, 15, 20, 30, 40, 50, 100, 150, 200, 250, 300, 2500),
                                 labels =c("<3", "3-5", "5-7", "7-10", "10-15", "15-20", "20-30", "30-40",
                                           "40-50", "50-100", "100-150", "150-200", "200-250", "250-300", ">300"),
                                  #breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                                  #labels <- c("<1", "1-3","1-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                                 include.lowest = TRUE))
sink(tempfile())  # Redirect console output to a temporary file
pred_2020 <- exact_extract(pred_inc_ras_2020, bnd, 'mean') 
sink()  # Reset console output
bnd_2020  <- bind_cols(bnd, pred_2020)
admin0 <- st_union(bnd_2020)

bnd2_2020 <- pivot_longer(bnd_2020, cols = mean.pred_01:mean.pred_12, names_to = 'month', values_to = 'inc') %>% 
  select(adm2 = NAME, month, inc) %>% 
  mutate(month = str_remove(month, 'mean.pred_') %>% as.numeric(),
         inc_cut = cut(inc, 
                       breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                       labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                        #breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                        #labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                       include.lowest = TRUE)) %>% 
    left_join(pop_dat) %>% 
  mutate(predicted_cases = inc*population/(1000*12)) %>% 
  select(adm1, adm2, month, predicted_cases, population, incidence_PMO = inc, inc_cut) #PMO:Person Month Observed

#write_csv(st_drop_geometry(bnd2) %>% select(-inc_cut), "/home/odiao/Plots_months_files/Output/predicted_incidence_monthly_2020.csv")
#st_write(bnd2 %>% select(-inc_cut), "/home/odiao/Plots_months_files/Output/predicted_incidence_monthly_2020.shp")

ggplot(bnd2_2020)+
  geom_sf(mapping = aes(fill = incidence_PMO))+ #inc: modified by OD
  facet_wrap(~month)+
  #scale_fill_viridis(option = "B", na.value = 'transparent', direction = -1)+
  scale_fill_distiller(palette = "RdYlGn")+
  theme_void()+
  labs(title = "Predicted Incidence in 2020", fill = "Cases per 1000 PMO")+
  theme(
    strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 16),               # Caption size
        legend.text = element_text(size = 16),                # Legend text size
        legend.title = element_text(size = 16, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  # Largeur des clés
        legend.key.height = unit(0.5, "cm") 
  )
#ggsave("06 Presentation materials/Maps/predicted_inc_district_viridis.png")
ggsave("/home/odiao/Plots_months_files/Output/2020_predicted_inc_monthly_district_viridis.png")
```




```{r fig-dev_2020, fig.width=14, fig.height=8}
pred_inc_monthly_2020 <- terra::as.data.frame(pred_inc_ras_2020, xy = TRUE) %>% 
  pivot_longer(cols = names(pred_inc_ras_2020), names_to = "month", values_to = "inc") %>% 
  mutate(
    month = str_remove(month, 'pred_') %>% as.numeric()
  ) %>%
  group_by(x, y) %>% 
  mutate(
    annual_avg = mean(inc, na.rm = TRUE),       # Calculate annual average per location
    monthly_dev = inc - annual_avg             # Calculate monthly deviations
  ) %>% 
  ungroup()                                    # Ungroup for further operations

pred_inc_monthly_2020 <- pred_inc_monthly_2020 %>% 
  mutate(
    inc_cut = cut(
      inc, 
      breaks <- c(-300, -200, -100, -50, 1, 50, 100, 300, 1200),
      labels <- c("< -200", "-200 to -100", "-100 to -50", "-50 to 1", "1 to 50", "50 to 100","100 to 300", "> 300"),
      include.lowest = TRUE
    )
  )

#Plot categorized incidence
ggplot(pred_inc_monthly_2020, aes(x = x, y = y, fill = inc_cut)) +
  geom_tile() +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  coord_equal() +
  theme_minimal() +
  labs(
    title = "Deviation from annual average (2020)",
    x = "Longitude",
    y = "Latitude",
    fill = "Deviation"
  ) +
  facet_wrap(~ month, ncol = 4)+
  theme(strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 12),               # Caption size
        axis.title.x = element_text(size = 16),  # X-axis label size
        axis.title.y = element_text(size = 16),  # Y-axis label size
        axis.text.x = element_text(size = 14),   # X-axis tick size
        axis.text.y = element_text(size = 14),   # Y-axis tick size
        legend.text = element_text(size = 14),                # Legend text size
        legend.title = element_text(size = 14, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  # Largeur des clés
        legend.key.height = unit(0.5, "cm") 
  )
ggsave("/home/odiao/Plots_months_files/Output/2020_deviation_inc_monthly_viridis.png")
```


### Monthly maps in 2021
```{r Incidence maps 2021, fig.width=14, fig.height=8}
pred_inc_ras_2021 <- rast(raster_files[13:24])
names(pred_inc_ras_2021) <- c("_2021_pred_01", "_2021_pred_02","_2021_pred_03","_2021_pred_04",
                         "_2021_pred_05","_2021_pred_06","_2021_pred_07","_2021_pred_08",
                         "_2021_pred_09","_2021_pred_10", "_2021_pred_11","_2021_pred_12")
#plot(pred_inc_ras)
#summary(pred_inc_ras) #This can help to define breaks

pred_inc_2021 <- terra::as.data.frame(pred_inc_ras_2021, xy=TRUE) %>% 
            pivot_longer(cols=names(pred_inc_ras_2021), names_to = "month", values_to = "inc") %>% 
            mutate(month = str_remove(month, '_2021_pred_'),
                   month = as.numeric(month)
                   ) %>% 
            mutate(inc_cut = cut(inc, 
                                 breaks = c(0,3, 5, 7, 10, 15, 20, 30, 40, 50, 100, 150, 200, 250, 300, 2500),
                                 labels =c("<3", "3-5", "5-7", "7-10", "10-15", "15-20", "20-30", "30-40",
                                           "40-50", "50-100", "100-150", "150-200", "200-250", "250-300", ">300"),
                                  #breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                                  #labels <- c("<1", "1-3","1-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                                 include.lowest = TRUE))
sink(tempfile())  # Redirect console output to a temporary file
pred_2021 <- exact_extract(pred_inc_ras_2021, bnd, 'mean') 
sink()  # Reset console output
bnd_2021 <- bind_cols(bnd, pred_2021)
admin0 <- st_union(bnd_2021)

bnd2_2021 <- pivot_longer(bnd_2021, cols =mean._2021_pred_01:mean._2021_pred_12 , names_to = 'month', values_to = 'inc') %>% 
  select(adm2 = NAME, month, inc) %>% 
  mutate(month = str_remove(month, 'mean._2021_pred_') %>% as.numeric(),
         inc_cut = cut(inc, 
                       breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                        labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                        #breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                        #labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                       include.lowest = TRUE)) %>% 
    left_join(pop_dat) %>% 
  mutate(predicted_cases = inc*population/(1000*12)) %>% 
  select(adm1, adm2, month, predicted_cases, population, incidence_PMO = inc, inc_cut) #PMO:Person Month Observed

#write_csv(st_drop_geometry(bnd2) %>% select(-inc_cut), "/home/odiao/Plots_months_files/Output/predicted_incidence_monthly_2021.csv")
#st_write(bnd2 %>% select(-inc_cut), "/home/odiao/Plots_months_files/Output/predicted_incidence_monthly_2021.shp")

ggplot(bnd2_2021)+
  geom_sf(mapping = aes(fill = incidence_PMO))+ #inc: modified by OD
  facet_wrap(~month)+
  #scale_fill_viridis(option = "B", na.value = 'transparent', direction = -1)+
  scale_fill_distiller(palette = "RdYlGn")+
  theme_void()+
  labs(title = "Predicted Incidence in 2021", fill = "Cases per 1000 PMO")+
  theme(
    strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 16),               # Caption size
        legend.text = element_text(size = 16),                # Legend text size
        legend.title = element_text(size = 16, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  # Largeur des clés
        legend.key.height = unit(0.5, "cm") 
  )
#ggsave("06 Presentation materials/Maps/predicted_inc_district_viridis.png")
ggsave("/home/odiao/Plots_months_files/Output/2021_predicted_inc_monthly_district_viridis.png")
```
 
```{r fig-dev_2021, fig.width=14, fig.height=8}
# Reshape and calculate the annual average
pred_inc_monthly_2021 <- terra::as.data.frame(pred_inc_ras_2021, xy = TRUE) %>% 
  pivot_longer(cols = names(pred_inc_ras_2021), names_to = "month", values_to = "inc") %>% 
  mutate(
    month = str_remove(month, '_2021_pred_') %>% as.numeric()
  ) %>%
  group_by(x, y) %>% 
  mutate(
    annual_avg = mean(inc, na.rm = TRUE),       # Calculate annual average per location
    monthly_dev = inc - annual_avg             # Calculate monthly deviations
  ) %>% 
  ungroup()                                    # Ungroup for further operations

pred_inc_monthly_2021 <- pred_inc_monthly_2021 %>% 
  mutate(
    inc_cut = cut(
      inc, 
      breaks <- c(-300, -200, -100, -50, 1, 50, 100, 300, 1200),
      labels <- c("< -200", "-200 to -100", "-100 to -50", "-50 to 1", "1 to 50", "50 to 100","100 to 300", "> 300"),
      include.lowest = TRUE
    )
  )

#Plot categorized incidence
ggplot(pred_inc_monthly_2021, aes(x = x, y = y, fill = inc_cut)) +
  geom_tile() +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  coord_equal() +
  theme_minimal() +
  labs(
    title = "Deviation from annual average (2021)",
    x = "Longitude",
    y = "Latitude",
    fill = "Deviation"
  ) +
  facet_wrap(~ month, ncol = 4)+
  theme(strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 12),               # Caption size
        axis.title.x = element_text(size = 16),  # X-axis label size
        axis.title.y = element_text(size = 16),  # Y-axis label size
        axis.text.x = element_text(size = 14),   # X-axis tick size
        axis.text.y = element_text(size = 14),   # Y-axis tick size
        legend.text = element_text(size = 14),                # Legend text size
        legend.title = element_text(size = 14, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  # Largeur des clés
        legend.key.height = unit(0.5, "cm") 
  )

ggsave("/home/odiao/Plots_months_files/Output/2021_deviation_inc_monthly_viridis.png")
```

### Monthly maps in 2022
```{r Incidence maps 2022, fig.width=14, fig.height=8}
pred_inc_ras_2022 <- rast(raster_files[25:36])
names(pred_inc_ras_2022) <- c("_2022_pred_01", "_2022_pred_02","_2022_pred_03","_2022_pred_04",
                         "_2022_pred_05","_2022_pred_06","_2022_pred_07","_2022_pred_08",
                         "_2022_pred_09","_2022_pred_10", "_2022_pred_11","_2022_pred_12")
#plot(pred_inc_ras)
#summary(pred_inc_ras) #This can help to define breaks

pred_inc_2022 <- terra::as.data.frame(pred_inc_ras_2022, xy=TRUE) %>% 
            pivot_longer(cols=names(pred_inc_ras_2022), names_to = "month", values_to = "inc") %>% 
            mutate(month = str_remove(month, '_2022_pred_'),
                   month = as.numeric(month)
                   ) %>% 
            mutate(inc_cut = cut(inc, 
                                 breaks = c(0,3, 5, 7, 10, 15, 20, 30, 40, 50, 100, 150, 200, 250, 300, 2500),
                                 labels =c("<3", "3-5", "5-7", "7-10", "10-15", "15-20", "20-30", "30-40",
                                           "40-50", "50-100", "100-150", "150-200", "200-250", "250-300", ">300"),
                                  #breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                                  #labels <- c("<1", "1-3","1-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                                 include.lowest = TRUE))
sink(tempfile())  # Redirect console output to a temporary file
pred_2022 <- exact_extract(pred_inc_ras_2022, bnd, 'mean') 
sink()  # Reset console output
bnd_2022 <- bind_cols(bnd, pred_2022)
admin0 <- st_union(bnd_2022)

bnd2_2022 <- pivot_longer(bnd_2022, cols = mean._2022_pred_01:mean._2022_pred_12, names_to = 'month', values_to = 'inc') %>% 
  select(adm2 = NAME, month, inc) %>% 
  mutate(month = str_remove(month, 'mean._2022_pred_') %>% as.numeric(),
         inc_cut = cut(inc, 
                       breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                        labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                        #breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                        #labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                       include.lowest = TRUE)) %>% 
    left_join(pop_dat) %>% 
  mutate(predicted_cases = inc*population/(1000*12)) %>% 
  select(adm1, adm2, month, predicted_cases, population, incidence_PMO = inc, inc_cut) #PMO:Person Month Observed

#write_csv(st_drop_geometry(bnd2) %>% select(-inc_cut), "/home/odiao/Plots_months_files/Output/predicted_incidence_monthly_2022.csv")
#st_write(bnd2 %>% select(-inc_cut), "/home/odiao/Plots_months_files/Output/predicted_incidence_monthly_2022.shp")

ggplot(bnd2_2022)+
  geom_sf(mapping = aes(fill = incidence_PMO))+ #inc: modified by OD
  facet_wrap(~month)+
  #scale_fill_viridis(option = "B", na.value = 'transparent', direction = -1)+
  scale_fill_distiller(palette = "RdYlGn")+
  theme_void()+
  labs(title = "Predicted Incidence in 2022", fill = "Cases per 1000 PMO")+
  theme(
    strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 16),               # Caption size
        legend.text = element_text(size = 16),                # Legend text size
        legend.title = element_text(size = 16, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  # Largeur des clés
        legend.key.height = unit(0.5, "cm") 
  )
#ggsave("06 Presentation materials/Maps/predicted_inc_district_viridis.png")
ggsave("/home/odiao/Plots_months_files/Output/2022_predicted_inc_monthly_district_viridis.png")
```


```{r fig-dev_2022, fig.width=14, fig.height=8}
# Reshape and calculate the annual average
pred_inc_monthly_2022 <- terra::as.data.frame(pred_inc_ras_2022, xy = TRUE) %>% 
  pivot_longer(cols = names(pred_inc_ras_2022), names_to = "month", values_to = "inc") %>% 
  mutate(
    month = str_remove(month, '_2022_pred_') %>% as.numeric()
  ) %>%
  group_by(x, y) %>% 
  mutate(
    annual_avg = mean(inc, na.rm = TRUE),       # Calculate annual average per location
    monthly_dev = inc - annual_avg             # Calculate monthly deviations
  ) %>% 
  ungroup()                                    # Ungroup for further operations

pred_inc_monthly_2022 <- pred_inc_monthly_2022 %>% 
  mutate(
    inc_cut = cut(
      inc, 
      breaks <- c(-300, -200, -100, -50, 1, 50, 100, 300, 1200),
      labels <- c("< -200", "-200 to -100", "-100 to -50", "-50 to 1", "1 to 50", "50 to 100","100 to 300", "> 300"),
      include.lowest = TRUE
    )
  )

#Plot categorized incidence
ggplot(pred_inc_monthly_2022, aes(x = x, y = y, fill = inc_cut)) +
  geom_tile() +
  scale_fill_brewer(palette = "RdYlBu", direction = -1) +
  coord_equal() +
  theme_minimal() +
  labs(
    title = "Deviation from annual average (2022)",
    x = "Longitude",
    y = "Latitude",
    fill = "Deviation"
  ) +
  facet_wrap(~ month, ncol = 4)+
theme(strip.text = element_text(size = 16, face = "bold"),
      plot.title = element_text(size = 16, face = "bold"),   # Title size
      plot.caption = element_text(size = 12),               # Caption size
      axis.title.x = element_text(size = 16),  # X-axis label size
      axis.title.y = element_text(size = 16),  # Y-axis label size
      axis.text.x = element_text(size = 14),   # X-axis tick size
      axis.text.y = element_text(size = 14),   # Y-axis tick size
      legend.text = element_text(size = 14),                # Legend text size
      legend.title = element_text(size = 14, face = "bold"), # Legend title size
      legend.position = 'bottom',
      legend.key.width = unit(2., "cm"),                  # Largeur des clés
      legend.key.height = unit(0.5, "cm") 
  )

#ggsave("06 Presentation materials/Maps/predicted_inc_district_viridis.png")
ggsave("/home/odiao/Plots_months_files/Output/2022_deviation_inc_monthly_district_viridis.png")
```


#### Data grouped by calendar month and calculate the median
```{r fig-group_month, fig.width=14, fig.height=8}
# Extract year and month from file names using regex
raster_info <- data.frame(
  file = raster_files,
  year = sub("^(\\d{4})\\.\\d{2}_posterior_median_cases_total_per_1000_PYO\\.tif$", "\\1", basename(raster_files)),
  month = sub("^\\d{4}\\.(\\d{2})_posterior_median_cases_total_per_1000_PYO\\.tif$", "\\1", basename(raster_files))
)

# Initialize a list to store monthly medians
monthly_medians <- list()

# Loop through each month
for (month in unique(raster_info$month)) {
  # Subset raster files for the current month
  monthly_files <- raster_info$file[raster_info$month == month]
  # Stack the rasters for the current month
  raster_stack <- stack(monthly_files)
  # Calculate the pixel-wise median
  monthly_median <- calc(raster_stack, median, na.rm = TRUE)
  # Store the result
  monthly_medians[[month]] <- monthly_median
}
# Combine all monthly medians into a single raster stack (optional)
all_monthly_medians <- stack(monthly_medians)


# Write output rasters 
#for (i in seq_along(monthly_medians)) {
 # writeRaster(monthly_medians[[i]], 
  #            filename = file.path("/home/odiao/Plots_months_files/Output/median_month_", names(monthly_medians)[i], #".tif"), 
 #             overwrite = TRUE)
#}

sink(tempfile())  # Redirect console output to a temporary file
pred_all_monthly_medians <- exact_extract(all_monthly_medians, bnd, 'mean') 
sink()  # Reset console output
bnd_all_monthly_medians <- bind_cols(bnd, pred_all_monthly_medians)
admin0 <- st_union(bnd)

bnd2_all_monthly_medians <- pivot_longer(bnd_all_monthly_medians, cols = names(pred_all_monthly_medians), names_to = 'month', values_to = 'inc') %>% 
  select(adm2 = NAME, month, inc) %>% 
  mutate(
    month = str_remove(month, 'mean.X'), 
    month = as.numeric(month),
         inc_cut = cut(inc, 
                        breaks = c(0, 1,3,5,10,50,100,200,300,1600),
                        labels <- c("<1", "1-3","3-5", "5-10", "10-50", "50-100", "100-200", "200-300", ">300"),
                       include.lowest = TRUE)) %>% 
    left_join(pop_dat) %>% 
  mutate(predicted_cases = inc*population/(1000*12)) %>% 
  select(adm1, adm2, month, predicted_cases, population, incidence_PMO = inc, inc_cut) #PMO:Person Month Observed
ggplot(bnd2_all_monthly_medians)+
  geom_sf(mapping = aes(fill = incidence_PMO))+ 
  facet_wrap(~month)+
  #scale_fill_viridis(option = "B", na.value = 'transparent', direction = -1)+
  scale_fill_distiller(palette = "RdYlGn")+
  theme_void()+
  labs(title = "Monthly Median of Incidence in 2020:2022", fill = "Cases per 1000 PMO")+
  theme(
    strip.text = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"),   # Title size
        plot.caption = element_text(size = 16),               # Caption size
        legend.text = element_text(size = 16),                # Legend text size
        legend.title = element_text(size = 16, face = "bold"), # Legend title size
        legend.position = 'bottom',
        legend.key.width = unit(2., "cm"),                  
        legend.key.height = unit(0.5, "cm") 
  )
ggsave("/home/odiao/Plots_months_files/Output/2020_2021_2022_median_inc_monthly_district_viridis.png")
```


# Conclusion
We have prodcuced monthly maps from 2020 to 2022. These maps show the variability and the intensity of malaria transmission by month where we clearly distinguish that the months from July to November have the high number of cases meaning that where the high transmission of malaria occurs.